---
title: "Synthetic populations from multi-study IPD with muvamere"
author: "Pete Dodd"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synthetic populations from multi-study IPD with muvamere}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## Stan fits take minutes, so they only run when asked for, e.g.
## Sys.setenv(MUVAMERE_VIGNETTE_FIT = "true") before building.
## TODO: or precompute the fit and ship it in inst/extdata
run_fits <- identical(Sys.getenv("MUVAMERE_VIGNETTE_FIT"), "true")
```

# Overview

**muvamere** (MUlti-VAriate MEta-REgression) fits hierarchical
multivariate regression models to individual patient data (IPD) pooled
from several studies, and uses the fit to generate synthetic ("artificial")
populations, e.g. for economic evaluation. The intended workflow is:

| Step | Function(s) |
|------|-------------|
| 1. Assemble `Y` (outcomes), `X` (covariates), `study` | your code |
| 2. Fit the hierarchical model | `mvn_infer_mlm_mixed()` |
|  (all-continuous data) | `mvn_infer_mlm_sparse()` |
| 3. Check convergence | `mvn_diagnose()`, `attr(fit, "diagnose")` |
| 4. Inspect global parameters | `mvn_extract_hyperparams()` |
| 5. Make a shareable, data-free object | `mvn_make_generator()` |
|  (audit it) | `mvn_check_generator()` |
| 6. Generate synthetic cohorts | `mvn_generate_AP()` |

Steps 2-5 can be done by whoever holds the data; the generator object
from step 5 contains no records and can be passed on for step 6.

TODO: a sentence or two of motivation / link to the paper.

# The model

For record $j$ in study $i$, with covariate row $x_{ij}$, the latent
outcome vector (one element per variate) is

$$
y_{ij} \sim \mathrm{MVN}\left(x_{ij} B_i,\;
  \mathrm{diag}(\tau_i)\, \Omega_i\, \mathrm{diag}(\tau_i)\right).
$$

* **Regression coefficients** are partially pooled across studies:
  $B_i = B_M + B_S \circ Z_i$, $Z_i \sim N(0, 1)$ elementwise.
* **Scales** follow a log-normal hierarchy,
  $\log \tau_{iv} \sim N(\mu_{\tau,v}, \sigma_{\tau,v})$
  (`ltaum`, `lsig`).
* **Correlations.** A global correlation matrix $\Omega_G$
  (`Omega_global`) has a regularized horseshoe prior on its off-diagonal
  elements (Piironen and Vehtari 2017), so most pairs are shrunk towards
  zero. Each study's correlations deviate from it,
  $\Omega_{i,kl} \sim N(\Omega_{G,kl}, \kappa)$, with the shared
  deviation strength $\kappa$ estimated; non positive-definite study
  matrices are rejected.
* **Binary variates** (in `mvn_infer_mlm_mixed()`) are the indicator that
  the latent variate exceeds 0 (a multivariate probit), with latent scale
  fixed at 1. Their correlations are therefore latent (tetrachoric or
  biserial) correlations. The likelihood is computed exactly with the GHK
  algorithm.
* **Missing values** in `Y` are handled under missing-at-random.

A new synthetic study draws its own $B$, $\tau$ and $\Omega$ from these
distributions, so generated cohorts carry the between-study heterogeneity
seen in the data.

TODO: say more on the choice of priors / point to model_formulations.

# A worked example with simulated data

```{r libs}
library(muvamere)
```

## Simulating data with a known truth

Four variates in an interleaved order: two continuous (`c1`, `c2`) and two
binary (`b1`, `b2`), with a covariate `x`. The true global correlation is
sparse: only `c1`-`b1` and `c2`-`b2` are correlated.

```{r truth}
set.seed(1)
vars <- c("c1", "b1", "c2", "b2")
binary <- c(FALSE, TRUE, FALSE, TRUE)
OG <- diag(4)
dimnames(OG) <- list(vars, vars)
OG["c1", "b1"] <- OG["b1", "c1"] <- 0.6
OG["c2", "b2"] <- OG["b2", "c2"] <- -0.4
## rows: intercept, x; columns: variates
betag <- matrix(c(1, 0.2, 1, -0.3, 0.5, 0.3, 1, 0.1), 2, 4,
  dimnames = list(c("int", "x"), vars)
)
betag
```

`mvn_sample_study_kappa()` simulates one study from the model given its
global parameters. Here we make three studies of 150 records:

```{r simulate}
S <- 3
Np <- 150
Xlist <- replicate(S, cbind(int = 1, x = runif(Np)), simplify = FALSE)
Ylist <- lapply(Xlist, function(X) {
  mvn_sample_study_kappa(X, betag,
    sigb = matrix(0.05, 2, 4), kappa = 0.05,
    ltaum = c(0, 0), lsig = c(0.1, 0.1), OmegaG = OG, binary = binary
  )
})
Y <- do.call(rbind, Ylist)
X <- do.call(rbind, Xlist)
study <- rep(seq_len(S), each = Np)
head(Y)
```

Add some scattered missingness to one continuous and one binary column:

```{r missing}
Y0 <- Y # complete copy, used for the single-study example below
Y[sample(nrow(Y), 40), "c2"] <- NA
Y[sample(nrow(Y), 40), "b2"] <- NA
colSums(is.na(Y))
```

## Fitting

`binary` can be a logical vector, column indices or column names. Records
can be in any order; they are sorted by study internally.

```{r fit, eval = run_fits}
fit <- mvn_infer_mlm_mixed(Y, X, study,
  binary = c("b1", "b2"),
  iter = 1000, chains = 2, cores = 2, seed = 3, refresh = 0
)
```

With all-continuous data (and no missing values) use
`mvn_infer_mlm_sparse()` instead, which takes the same arguments apart
from `binary` and `min_joint` and is much faster.

Main tuning arguments (defaults were used throughout validation):

* `p0`: prior guess for the number of non-zero correlations among the
  `choose(NV, 2)` pairs (default 10%, rounded up);
* `slab_scale`, `slab_df`: regularize large correlations;
* `kappa_prior_scale`: half-normal scale for $\kappa$;
* `betaM_prior_sd`, `betaS_prior_sd`, `tauM_prior_sd`, `tauS_prior_sd`.

Anything else in `...` goes to `rstan::sampling()`.

## Checking convergence

Every fit is passed through `mvn_diagnose()`, stored as an attribute, and a
warning is given if sampling is severely unstable.

```{r diagnose, eval = run_fits}
attr(fit, "diagnose")
```

`mvn_diagnose(fit, ...)` can be rerun with other thresholds. A handful of
divergences can occur with weak data (small studies, large $\kappa$).

TODO: advice on what to do if the checks fail (more iter, adapt_delta,...).
TODO: this toy example (3 studies x 150) gave 129/1000 divergences and
ESS 6.8% when test-rendered on 2026-09-26 - pick a better-behaved example
(more/larger studies?) or discuss it here.

## Global parameters

`mvn_extract_hyperparams()` returns posterior means of the global
parameters, named and in the original column order (internally the model
puts continuous variates first):

```{r hyper, eval = run_fits}
h <- mvn_extract_hyperparams(fit)
round(h$OmegaG, 2)
round(h$betag, 2)
h$kappa
```

Compare with the truth:

```{r compare, eval = run_fits}
mvn_compare_cov(OG, h$OmegaG)$rmse[1]
mvn_compare_means(betag, h$betag)$rmse[1]
```

For full posterior summaries use the stanfit directly, e.g.
`rstan::summary(fit, pars = "Omega_global")`.

## A data-free generator

A stanfit can hold record-level information (here, per-record GHK
uniforms if saved). `mvn_make_generator()` keeps only posterior draws of
the global hyperparameters plus variable metadata:

```{r generator, eval = run_fits}
gen <- mvn_make_generator(fit, ndraws = 500)
gen
## audit, including for anything with a record-sized dimension
mvn_check_generator(gen, n_records = nrow(Y))
```

It can be saved and shared where the data cannot:

```{r save, eval = FALSE}
saveRDS(gen, "my_generator.rds")
gen <- readRDS("my_generator.rds")
mvn_check_generator(gen) # re-audit anything received from elsewhere
```

Note this removes the data, not all information derived from it: the
hyperparameter draws are aggregate summaries of the fitted studies, and
this is not a formal privacy guarantee.

## Generating synthetic cohorts

`mvn_generate_AP()` simulates one new study per covariate matrix in
`Xlist` (same columns as the fitted `X`). Given a generator, each
synthetic study uses a different posterior draw, propagating parameter
uncertainty; given a stanfit, posterior means are plugged in.

```{r generate, eval = run_fits}
Xnew <- replicate(5, cbind(int = 1, x = runif(1000)), simplify = FALSE)
ap <- mvn_generate_AP(gen, Xnew)
head(ap)
## binaries come back as 0/1
aggregate(cbind(c1, b1, c2, b2) ~ studyno, data = ap, FUN = mean)
```

TODO: a plot comparing observed vs synthetic marginals/correlations.

# Single-study fits

For a single study, `mvn_infer()` fits a (continuous) multivariate
regression with an LKJ prior on the correlation. Passing new covariates
`Z` gives posterior-predictive draws, extracted as a
`[draw, row of Z, variate]` array:

```{r single, eval = run_fits}
fit1 <- mvn_infer(Y0[study == 1, c("c1", "c2")], X[study == 1, ],
  Z = cbind(1, c(0, 0.5, 1)), iter = 1000, chains = 2, cores = 2,
  refresh = 0
)
pred <- mvn_extract_predictions(fit1)
dim(pred)
```

`mvn_infer()` does not handle missing values, hence the complete `Y0`.

# Practical notes and limitations

* **Variable types.** Continuous and binary only; ordinal/categorical
  variates are not supported. Binary columns must be 0/1 (or logical).
* **Missing data.** Allowed in `Y` (mixed model only), not in `X`. Every
  variable and every pair of variables must be jointly observed in at
  least `min_joint` (default 5) records in *every* study; a variable
  absent (or almost absent) from a whole study gives an error - drop the
  variable or the study. A binary that is constant within a study gives
  a warning.
* **Categorical covariates** (e.g. HIV status) go into `X` as dummies.
* **Run time.** The mixed model evaluates the likelihood record by record:
  roughly an hour for 2 chains $\times$ 1000 iterations with 10 variates
  and ~2200 records on a laptop; much longer at 20 variates.
* **Privacy.** Share generators, not fits.

TODO: anything on choosing the covariates / target-population `X`.

# References

Piironen J, Vehtari A (2017). Sparsity information and regularization in
the horseshoe and other shrinkage priors. *Electronic Journal of
Statistics* 11(2):5018-5051. doi:10.1214/17-EJS1337SI

TODO: the muvamere / DTB_AP paper.
